---
title: Tweak printCoefmat() - better Round()ing / Zapping Zeros
description: "In rare cases, the rounding, e.g. of the std.err. column in
              printCoefmat() is undesirably rounding too much and then
			  depicting inaccurate values. This is from the recent R-devel (mailing
			  list), thread starting at
			  https://stat.ethz.ch/pipermail/r-help/2023-July/477688.html"
author: Martin Maechler
output: html_document
categories: [I/O, Wishlist] # from https://github.com/r-devel/r-project-sprint-2023/labels
comments:
  giscus: 
    repo: "r-devel/r-project-sprint-2023"
    repo-id: "R_kgDOIhAibA"
    category: "Proposals"
    category-id: "DIC_kwDOIhAibM4CW3GY"
    mapping: "title"
    reactions-enabled: true
    loading: lazy
---

```{=html}
<!-- 
This template is based on https://contributor.r-project.org/r-project-sprint-2023/projects/quartz-alpha-mask/.
The sections are provided as a guide and may not be appropriate to your proposal, feel free to skip or change sections.

Please label any R code chunks, especially those producing images.
-->
```
```{r setup}

```

## Background

Printing of e.g.  `lm()` results; but also glm(), anova(), etc; not just
package `{stats}` but used by other packages as well, e.g., `lme4`.

## Problem statement

```{r lm-example}
## Slightly modified from Shu Fai Cheung's example on R-devel (the mailing list)
set.seed(123)
n <- 1000
x1 <- rnorm(n)
x2 <- rnorm(n)
y <- 0.000754 + .5 * x1 + .6 * x2 + .036*rnorm(n)
dat <- data.frame(x1, x2, y)
fm <- lm(y ~ x1 + x2, dat)
##    ---
sfm <- summary(fm)
print(sfm)# "fine" - but want intercept "zapped to zero" ==>
print(sfm, zap.ind=1)# now, get two extra  trailing '00'  for the coefficient column
## a "bug"
```
........
........

Coefficients:
            Estimate Std. Error t value Pr(>|t|)    
(Intercept) 0.000000   0.001115     0.0        1    
x1          0.499200   0.001128   442.4   <2e-16 ***
x2          0.601000   0.001108   542.3   <2e-16 ***

........
.........
```
Here, we'd rather like to see correctly rounded coefficients
```
            Estimate Std. Error t value Pr(>|t|)    
(Intercept) 0.000000   0.001115     0.0        1    
x1          0.499263   0.001128   442.4   <2e-16 ***
x2          0.600990   0.001108   542.3   <2e-16 ***
```


## Proposed solution

Improve `printCoefmat()`.

The use of `zapsmall()` there, can round too much in these cases.

One option is to extend the meaning of the `zap.ind` argument and allow
*negative* indices to mean that the zapping should *not* be done with
`zapsmall()` but with a (to be defined inside `printCoefmat()`)
```{r zapOnlysmall-def}
zapOnlysmall <- function(x, dig) {
    x[abs(x) <= 10^-dig] <- 0
    x
}
```

I had tried the following patch

```
--- anova.R	(revision 84998)
+++ anova.R	(working copy)
@@ -105,7 +105,13 @@
 
     ok <- !(ina <- is.na(xm))
     ## zap before deciding any formats
-    for (i in zap.ind) xm[, i] <- zapsmall(xm[, i], digits)
+    zapOnlysmall <- function(x, dig) { x[abs(x) <= 10^-dig] <- 0 ; x }
+    for (i in zap.ind) { ai <- abs(i) # i < 0 [zapOnly..]  vs  i > 0 [zapsmall]:
+        xm[, ai] <- (if(i < 0) zapOnlysmall else zapsmall)(xm[, ai], digits)
+        ## problem: if one of these is a cs.ind (below) and is of larger order of magn., it is
+        ## -------  rounded too much here --> shows trailing 00 afterwards.
+	## ==> TODO: detect and *fix* rather than needing negative zap.ind for zapOnlysmall
+    }
```

which "solved" the problem, but actually also changes other examples
(not shown here) where a change is *not* necessary.

Also, see my "TODO" above, I think we could try to make this automatic, so
the  `zap.ind = 1` example above would just do "the right thing".


## Testing

This is not yet a strict test,
but code you can/should use to get a broader picture.
Note that this produces a few _pages_ of R console output :

```{r loop-many-digs-scales}
cf <- coef(sfm)

for(dig in c(3:5, 7, 10)) {
    cat("\ndigits = ", dig,":\n================\n")
    for(f in 10^c(-6, -3:3, 6, 9)) {
        cat("cf[,1:2] := cf0[,1:2] * f, f=", formatC(f),":\n")
        fcf <- cf; fcf[,1:2] <- f*cf[,1:2]
        printCoefmat(fcf, signif.legend=FALSE, digits=dig) # using e-07 for intercept --> everything in sci.format
        cat(" ... and with zap.ind = 1:\n")
        printCoefmat(fcf, signif.legend=FALSE, digits=dig, zap.ind = 1)
        cat("-------------------------\n\n")
    }
}
```

## Project requirements

Almost none.  Some understanding about floating point arithmetic, notably rounding.

## Project resources

## Project outcomes

A patch to the R source.  Code changes only in the first file (`anova.R`)
but a really useful patch also updates the documentation (`*.Rd`) notably
adding an example there.

As I have seen (with my own first fix), that it may change output of other
examples, and for some of these, the `tests/Examples/<pkg>-Ex.Rout.save`
files also need patching ( `tests/Examples/stats-Ex.Rout.save` will need an
update in any case *if* we modify/add examples on the help page:

```{sh files-to-change}
doc/NEWS.Rd
src/library/stats/R/anova.R
src/library/stats/man/printCoefmat.Rd
tests/Examples/{stats,graphics,datasets}-Ex.Rout.save
```


## Reactions and comments

```{=html}
<!-- 
Please leave the Reactions and comments section
- a Giscus comment box will be automatically added here 
-->
```
