---
title: Enhancing `sample.int` for unequal probability sampling
description: "" # Optional short description for post on Discussions thread
author: Ahmadou Dicko
output: html_document
categories: [C, R, Wishlist, Models]
comments:
  giscus:
    repo: "r-devel/r-project-sprint-2023"
    repo-id: "R_kgDOIhAibA"
    category: "Proposals"
    category-id: "DIC_kwDOIhAibM4CW3GY"
    mapping: "title"
    reactions-enabled: true
    loading: lazy
bibliography: ref.bib
---

## Problem statement

The method of unequal probability sampling without replacement, as implemented in `base::sample` (`base::sample.int`), relies on a sequential algorithm (coded in `C`). This algorithm does not respect the prescribed inclusion probabilities (as defined by the `prob` parameter). Consequently, it can produce a biased Horvitz-Thompson estimate.

The issue can affect other packages as illustrated by this [dplyr issue affecting slice_sample](https://github.com/tidyverse/dplyr/issues/6848).

To better understand the problem, consider the illustration below, following [@tille2023remarks].

```{r}
set.seed(123)
N <- 12
n <- 4
p <- (1:N)/sum(1:N)
pik <- n * p
pikest_base <- vector(mode = "numeric", length = N)
nsim <- 1e4
for (j in 1:nsim) {
  s <- sample.int(n = N, size = n, replace = FALSE, prob = p)
  pikest_base[s] <- pikest_base[s] + 1
}
pikest_base <- pikest_base/nsim
```

After estimating the inclusion probabilities with `sample.int`, we can compare them with algorithms from the `sampling` package, as highlighted in its documentation.

```{r}
library(sampling)
set.seed(42)
pikest_maxent <- sapply(1:nsim, \(i) UPmaxentropy(pik))
pikest_maxent <- rowMeans(pikest_maxent)

pikest_pivot <- sapply(1:nsim, \(i) UPpivotal(pik))
pikest_pivot <- rowMeans(pikest_pivot)

cbind(pik,
      pikest_base,
      pikest_maxent,
      pikest_pivot)
```

To evaluate the accuracy of the sampling algorithm, we can compute the following test statistic:


$$
z_k = \dfrac{(\hat{\pi_k} - \pi_k) \sqrt{M}}{\sqrt{\pi_k (1 - \pi_k)}}
$$

where $M$ is the number of simulations.

$z_k$ should be asymptotically normal under the null hypothesis that the sampling algorithm is correct. The implementation is as follows:

```{r}
test_stat <- function(est, pik)
  (est - pik) / sqrt(pik * (1 - pik)/nsim)
```

We can compute this test statistics for each method and check its normality using a `qqplot`.

```{r, echo = FALSE}
par(mfrow = c(2, 2))
set.seed(3)
theo <- rnorm(n = N)
qqnorm(theo, main = "Normal distribution (benchmark)")
qqline(theo)
qqnorm(test_stat(pikest_base, pik), main = "base::sample")
qqline(theo)
qqnorm(test_stat(pikest_maxent, pik), main = "sampling::UPmaxentropy")
qqline(theo)
qqnorm(test_stat(pikest_pivot, pik), main = "sampling::UPpivot")
qqline(theo)
```

These charts show how `base::sample` deviates more from normality compared to the other two competing algorithms available in the `sampling` package.


## Proposed solution

- **Option 1:**
  - Refine the documentation for `base::sample` and `base::sample.int`.
  - Suggest adding a concise paragraph in the `Details` section, explaining why the function is not suitable for unequal probability sampling based on inclusion probabilities.

- **Option 2:**
  - Introduce a new `C` function and an additional argument to `sample.int` to incorporate the new method.
  - Retain the original [`ProbSampleNoReplace` `C` function](https://github.com/wch/r-source/blob/7ad1c5790841572c88955b1da805d28174ff0b56/src/main/random.c#L406).
  - By default, `base::sample.int` should utilize `ProbSampleNoReplace` for unequal probability sampling without replacement. This ensures backward compatibility and prevents disruptions to existing code.


It is hard to pick one algorithm over the multitude of available unequal probability sampling algorithms. In theory, there's no best algorithm for unequal probability. However, the Maximum Entropy Sampling algorithm (`sampling::UPmaxentropy`) also known as Conditional Poisson Sampling is an algorithm with good properties (maximizing entropy) that can be used as a reference.

## Project requirements

- Good understanding of survey sampling algorithms.
- Good C and R coding skills.

## Project outcomes

An enhanced version of `sample` (`sample.int`) with a new argument to select the new algorithm for unequal probability sampling.

## References

<div id="refs"></div>
